---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library load, warning=FALSE, message=FALSE}
library(tidyverse)
```

## Loading and preprocessing the data
I load the data into a variable named proj_data, and check the percent of NAs in the steps variable.

```{r data load, warning=FALSE, cache=TRUE}
proj_data <- read_csv("./data/activity.csv")
head(proj_data)
class(proj_data)
mean(is.na(proj_data$steps))
```

The data is in tibble format, which is the format I need to use the various functions available from tidyverse.

```{r data time range, warning=FALSE}
interval_whole <-strptime(str_pad(proj_data$interval,4,pad="0"), format='%M %S')
proj_data_mutated <- mutate(proj_data,date_time = 
                              as.POSIXct(paste(proj_data$date, substr(interval_whole,15,20))))
head(proj_data_mutated)
nrow(proj_data_mutated)
proj_data_mutated_na_removed <- na.omit(proj_data_mutated)
head(proj_data_mutated_na_removed)


```

The percent of NA rows matches the difference in the original data and the data with NA values removed.

```{r data na check, warning=FALSE}
nrow(proj_data_mutated)
nrow(proj_data_mutated_na_removed)
(1-(15264/17568))
```

ggplot automatically removes NAs. Here, I plot the distribution minus NAs, and it's exactly the same as the steps distribution with NAs.

```{r data time plot no na, warning=FALSE}
ggplot(data = proj_data_mutated_na_removed)+geom_path(mapping = aes(x=date_time, y=steps))+
  labs(x="Time", y = "Steps")+
  ggtitle("Steps across time")
```

ggplot automatically removes NAs. Here, I plot the distribution minus NAs, and it's exactly the same as the steps distribution with NAs. 

```{r data-total-daily-steps, warning=FALSE}
proj_data_day_sum <- proj_data %>% group_by(date) %>%
      summarize(Sum.Steps = sum(steps))
ggplot(data = proj_data_day_sum, aes(x = date, y = Sum.Steps))+ 
  geom_bar(stat = "identity")+
  labs(x="Date", y = "Sum of Daily Steps")+
  ggtitle("Total Daily Steps")
```

The mean, excluding NAs, of the total steps per day is 10766.19.

```{r data-mean-daily-steps, warning=FALSE}
mean(na.omit(proj_data_day_sum$Sum.Steps))
```

## What is mean total number of steps taken per day?

Here is the distribution of the mean number of steps taken per day.

```{r data mean, warning=FALSE}
proj_data_mean <- proj_data_mutated_na_removed %>% group_by(date) %>%
                                        summarize(mean = mean(steps))
ggplot(data = proj_data_mean)+geom_col(mapping = aes(x=date, y=mean))+
  labs(x="Time", y = "Mean Daily Steps")+
  ggtitle("Mean Daily Steps")
```

Here is the distribution of the median number of steps taken per day for days with non-zero steps. The median for all days, including days with zero steps is zero for all days.

```{r data median, warning=FALSE}
proj_data_median <- proj_data_mutated_na_removed %>% group_by(date) %>%
  summarize(median = median(steps[steps > 0]))
ggplot(data = proj_data_median)+geom_col(mapping = aes(x=date, y=median))+
  labs(x="Time", y = "Median Daily Steps")+
  ggtitle("Median Daily Steps
          \n (excluding intervals with steps = 0")
```

## What is the average daily activity pattern?

As part of the exploration to answer this question, I have plotted the steps activity daily. 

```{r data daily average, warning=FALSE}
ggplot(data = proj_data_mutated)+
  geom_line(mapping = aes(x = interval, y = steps))+
  facet_wrap(~ date, nrow = 6)
```

Here is the mean number of steps by 5 second interval for all days.

```{r data daily average by time interval, warning=FALSE}
proj_data_interval <- proj_data_mutated_na_removed %>% group_by(interval) %>%
  summarize(mean = mean(steps))
ggplot(data = proj_data_interval)+geom_line(mapping = aes(x=interval, y=mean))+
  labs(x="Time Interval", y = "Mean Daily Steps")+
  ggtitle("Mean Daily Steps by Time Interval
          \n in Seconds for Entire Day")
```

The 5-minute interval that, on average, contains the maximum number of steps is found at 8 minutes and 35 seconds from the beginning of the day. The mean steps for this maximum interval is 206.

```{r max time interval, warning=FALSE}
filter(proj_data_interval, mean == max(mean))
```

## Imputing missing values
## Are there differences in activity patterns between weekdays and weekends?

I will answer the last two questions together. Actually, I want to answer them in the opposite order: I want to know the variations by day of the week and by separating the weekdays from the weekends, and then I want to carry out my imputation.

```{r factors, warning=FALSE}
proj_data_daysofweek <- mutate(proj_data_mutated,
  Day.of.Week = weekdays(as.POSIXct(date)))
Weekday <- c("Monday","Tuesday","Wednesday","Thursday","Friday")
Weekend <- c("Saturday","Sunday")
weekdays_weekend <- list(Weekday, Weekend)
factor(weekdays_weekend, levels = list(Weekday, Weekend))
proj_data_weekdays_only <- filter(proj_data_daysofweek,Day.of.Week %in% weekdays_weekend[[1]])
proj_data_weekdays_only <- mutate(proj_data_weekdays_only, Day.Type = c("Weekday"))
proj_data_weekend_only <- filter(proj_data_daysofweek,Day.of.Week %in% weekdays_weekend[[2]])
proj_data_weekend_only <- mutate(proj_data_weekend_only, Day.Type = c("Weekend"))
proj_data_factors <- rbind(proj_data_weekdays_only, proj_data_weekend_only)
head(proj_data_factors)
```

Intervals by day of the week.

```{r data-weekday-daily-average, warning=FALSE}
ggplot(data = proj_data_factors)+
  geom_line(mapping = aes(x = interval, y = steps))+
  facet_wrap(~ Day.of.Week, nrow = 2)
```

Intervals by if the day is a weekday or a weekend.

```{r data-weekend-daily-average, warning=FALSE}
ggplot(data = proj_data_factors)+
  geom_line(mapping = aes(x = interval, y = steps))+
  facet_wrap(~ Day.Type, nrow = 1)
```

In order to impute NAs, I follow this procedure:

1. I separate proj_data_factors, which is the data I have been plotting, which includes NAs into two datasets: one with NAs, one without. 

```{r step 1 split, warning=FALSE}
proj_data_factors_NA <- filter(proj_data_factors, 
                        is.na(proj_data_factors$steps) == TRUE)
nrow(proj_data_factors_NA)
proj_data_factors_notNA <- filter(proj_data_factors, 
                               is.na(proj_data_factors$steps) == FALSE)
nrow(proj_data_factors_notNA)
head(proj_data_factors_notNA)
```


2. I then split the data that has no NA values into weekday data and weekend data.

```{r step 2 split, warning=FALSE}
proj_data_factors_notNA_wday <- filter(proj_data_factors_notNA, Day.Type == "Weekday")
proj_data_factors_notNA_wend <- filter(proj_data_factors_notNA, Day.Type == "Weekend")                            
```

3. I calculate the mean steps for each time interval for the "no NA values" data separately for  weekday data and weekend data.

Here is the imputation distribution for weekdays.

```{r step-3-imp-dist-weekdays, warning=FALSE}
proj_data_notNA_wday_intervals <- proj_data_factors_notNA_wday %>% group_by(interval) %>%
                summarize(Mean.Interval = mean(steps))
ggplot(data=proj_data_notNA_wday_intervals)+geom_col(mapping = aes(x=interval, y=Mean.Interval))
```

Here is the imputation distribution for weekends.

```{r step-3-imp-dist-weekends, warning=FALSE}
proj_data_notNA_wend_intervals <- proj_data_factors_notNA_wend %>% group_by(interval) %>%
  summarize(Mean.Interval = mean(steps))
ggplot(data=proj_data_notNA_wend_intervals)+geom_col(mapping = aes(x=interval, y=Mean.Interval))
```

4. I use these distributions as the data I will impute. I assign the mean steps by interval from the "no NA values" weekday data to the weekday data with NA values for steps. I do the same for the weekend data by imputing weeking missing values with weekend modeled values from the mean steps in the "no NA values" weekend data.

```{r step-4-imputation, warning=FALSE}
proj_data_factors_NA_wday <- filter(proj_data_factors_NA, Day.Type == "Weekday")
proj_data_factors_NA_wend <- filter(proj_data_factors_NA, Day.Type == "Weekend")

for (i in 1:seq_along(proj_data_notNA_wday_intervals$interval)) {
    proj_data_factors_NA_wday$steps <- proj_data_notNA_wday_intervals$Mean.Interval
}

for (i in 1:seq_along(proj_data_notNA_wend_intervals$interval)) {
  proj_data_factors_NA_wend$steps <- proj_data_notNA_wend_intervals$Mean.Interval
}

proj_data_final_imputed <- rbind(proj_data_factors_notNA,proj_data_factors_NA_wday,
                                 proj_data_factors_NA_wend)
```

5. I perform checks on the data to make sure the imputation makes sense. 

```{r step-5-imputation check part 1, warning=FALSE}
proj_data_factors_NA_wday
proj_data_factors_NA_wend
```

```{r imputation-check-weekdays, warning=FALSE}
ggplot(data=proj_data_factors_NA_wday)+geom_col(mapping = aes(x=interval, y=steps))
```

```{r imputation-check-weekend, warning=FALSE}
ggplot(data=proj_data_factors_NA_wend)+geom_col(mapping = aes(x=interval, y=steps))
```

I check that the total rows in the final, imputed data, is the same as the starting data.

```{r step-5-imputation check part 2, warning=FALSE}
nrow(proj_data)
nrow(proj_data_final_imputed)
```

I check the intervals by day of the week for the imputed data.

```{r final-imputation-check-weekdays, warning=FALSE}
ggplot(data = proj_data_final_imputed)+
  geom_line(mapping = aes(x = interval, y = steps))+
  facet_wrap(~ Day.of.Week, nrow = 2)
```

I check the intervals by if the day is a weekday or a weekend for the imputed data.

```{r final-imputation-check-weekends, warning=FALSE}
ggplot(data = proj_data_final_imputed)+
  geom_line(mapping = aes(x = interval, y = steps))+
  facet_wrap(~ Day.Type, nrow = 1)
```

Finally, I draw the day by day plot and see that NA weekdays have been imputed and that NA weekends have been imputed. The means for the imputations are lower than the typical day because there are days with complete data (no NAs) with very small step counts, such as Oct 2, and Nov 15. In a more sophisticated imputation, I would have excluded those two days from my reference distribution to use as the data to assign to the NA days.

```{r final-daily-imputation-check, warning=FALSE}
baseplot <- ggplot(data = proj_data_final_imputed)
baseplot +  geom_line(mapping = aes(x = interval, y = steps))+
  facet_wrap(~ date, nrow = 6)
```

