---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library load, warning=FALSE, message=FALSE}
library(tidyverse)
```

## Loading and preprocessing the data
I load the data into a variable named proj_data, and check the percent of NAs in the steps variable.

```{r data load, warning=FALSE, cache=TRUE}
proj_data <- read_csv("./data/activity.csv")
head(proj_data)
class(proj_data)
mean(is.na(proj_data$steps))
```

The data is in tibble format, which is the format I need to use the various functions available from tidyverse.

```{r data time range, warning=FALSE}
interval_whole <-strptime(str_pad(proj_data$interval,4,pad="0"), format='%M %S')
proj_data_mutated <- mutate(proj_data,date_time = 
                              as.POSIXct(paste(proj_data$date, substr(interval_whole,15,20))))
head(proj_data_mutated)
nrow(proj_data_mutated)
proj_data_mutated_na_removed <- na.omit(proj_data_mutated)
head(proj_data_mutated_na_removed)


```

The percent of NA rows matches the difference in the original data and the data with NA values removed.

```{r data na check, warning=FALSE}
nrow(proj_data_mutated)
nrow(proj_data_mutated_na_removed)
(1-(15264/17568))
```

Steps distribution with NAs.

```{r data time plot, warning=FALSE}
ggplot(data = proj_data_mutated)+geom_path(mapping = aes(x=date_time, y=steps))+
  labs(x="Time", y = "Steps")+
  ggtitle("Steps across time")
```

ggplot utomatically removes NAs. Here, I plot the distribution minus NAs, and it's exactly the same as the steps distribution with NAs.


```{r data time plot no na, warning=FALSE}
ggplot(data = proj_data_mutated_na_removed)+geom_path(mapping = aes(x=date_time, y=steps))+
  labs(x="Time", y = "Steps")+
  ggtitle("Steps across time")
```

## What is mean total number of steps taken per day?





## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
